Election 2024

### 2024-10-31 paper/paper.qmd
add results and edi some of the quato code

@@

The structure of the paper is as follows

@@

# 2.1 Measurement 
The voter support data used in this analysis comes from raw polling information sourced from the Project 538 online database [@FiveThirtyEight]. This dataset reflects potential outcomes for the 2024 U.S. Presidential Election by aggregating various polls that capture public sentiment towards candidates Donald Trump and Kamala Harris. Each entry in the dataset corresponds to a specific poll conducted by different polling organizations, which measure the percentage of respondents expressing their support for each candidate in their respective state.

It’s important to note that each pollster employs unique sampling techniques and methodologies, which influence their numeric grades and transparency scores. To ensure the integrity of our analysis, we selected only those polls that originate from reputable pollsters, specifically targeting those with a numeric grade of 2.5 or higher. This approach is used so that the data reflects a robust and credible representation of voter preferences.

The polling data is collected through carefully structured survey questions, designed to elicit clear responses regarding voter preferences. By aggregating these individual responses, we are able to derive state-level averages for both candidates. This transformation of general voter sentiment into specific data points enables us to analyze the competitive landscape between Trump and Harris across swing states, providing a clearer understanding of electoral dynamics as they relate to the upcoming election.

@@

 # 2.2 Variables of Interest

@@

 # 4 Results {#sec-results}
The results of our analysis are as follows. We have Donald Trump predicted to win Arizona, Georgia and Michigan for a total of 42 of the 93 swing state votes. While Kamala Harris is projected to win Nevada, North Carolina, Pennsylvania and Wisconsin for the remaining 51 swing state votes. This results in a win for the democrats and the first female US president!

@@

The survey can be found here:
https://docs.google.com/forms/d/e/1FAIpQLSc5LYCAd0OmiIc-LHRFUVgVGuSMKzayVWZB7VOrsrDOe5742w/viewform?vc=0&c=0&w=1&flr=0
# References

>>>>

```{python}
#| echo: false
import pandas as pd
import matplotlib.pyplot as plt
import jupyter
from tabulate import tabulate
import numpy as np
```

@@

The structure of the paper is as follows: Section 2 describes the data collection process, detailing the sources, variables of interest, and criteria for selecting high-quality polls. Section 3 introduces the logistic regression model used for prediction, explaining the variables, weights, and methodology applied to forecast the election outcome. In Section 4, we present and analyze the model’s predictions, highlighting the probabilities of candidate success across key states. Section 5 discusses the model’s implications, limitations, and the observed contrasts between polling data.

@@

# 2.1 Measurement

The voter support data used in this analysis comes from raw polling information sourced from the Project 538 online database [@FiveThirtyEight]. This dataset provides predictions for candidate support for the 2024 U.S. Presidential Election by aggregating various polls that capture public sentiment towards candidates Donald Trump and Kamala Harris. Each entry in the dataset corresponds to a specific question around voting preference in a poll conducted by different polling organisations, which measure the percentage of respondents expressing their support for each candidate in their respective state. the surveys utilized one or multiple of the following method for rsponse gathering: Online Ad, Online Panel, Live Phone, text-to-web, mail-to-web, email, Probability panel, IVR. The diversity of the methods of collection supplies for a colorful array of polls capturing all facets of voters.

The polling data is collected through carefully structured survey questions, designed to elicit clear responses regarding voter preferences. By aggregating these individual responses, we are able to derive state-level sentiment for both candidates. This transformation of general voter sentiment into specific data points enables us to analyse the competitive landscape between Trump and Harris across swing states, providing a clearer understanding of electoral dynamics as they relate to the upcoming election.

@@

A swing state in the context of a US election is defined as "a state where the number of Democratic and Republican voters is about the same, that has an important influence on the result of the election of the United States President". States that do not fall under his definition are not likely to "swing" to another party as we approach the election, and thus their outcome can be reasonably assumed. Thus the election result is normally decided by which party wins just a handful of states in the electoral college system. By concentrating on these swing states, we aim to capture the most critical and uncertain areas of the electoral map, where voter preferences are most likely to sway the final result. Thus our analysis restricts the datasets to state level polls in the following states: Arizona, Pennsylvania, North Carolina, Georgia, Nevada, Michigan, and Wisconsin [@nyt_swing, @538]. In figure 3 we see the ammount that each party hold according to those sources, with the remaining swing states, seen in the middle of this graph set to determine if either Harris or Trump recieve over 270th electoral votes.

```{python}
# Define states and electoral votes, ensuring votes are listed in reverse for clarity in plot
states = ["California", "Connecticut", "Delaware", "Hawaii", "Illinois", "Maine", 
          "Maryland", "Massachusetts", "Minnesota", "New Jersey", "New York", "Oregon", 
          "Rhode Island", "Vermont", "Washington", "District of Columbia", "New Mexico", 
          "New Hampshire", "Colorado", "Virginia", "Arizona", "Georgia", "Michigan", 
          "Nevada", "North Carolina", "Pennsylvania", "Wisconsin", "Alabama", "Alaska", 
          "Arkansas", "Florida", "Idaho", "Indiana", "Iowa", "Kansas", "Kentucky", 
          "Louisiana", "Mississippi", "Missouri", "Montana", "Nebraska", "North Dakota", 
          "Ohio", "Oklahoma", "South Carolina", "South Dakota", "Tennessee", "Texas", 
          "Utah", "West Virginia", "Wyoming"]
votes = [54, 7, 3, 4, 19, 4, 10, 11, 10, 14, 28, 8, 4, 3, 12, 3, 5, 4, 10, 13, 
         11, 16, 15, 6, 16, 19, 10, 9, 3, 6, 30, 4, 11, 6, 6, 8, 8, 6, 10, 4, 5, 
         3, 17, 7, 9, 3, 11, 40, 6, 4, 3]

# Calculate cumulative votes
cumulative_votes = np.cumsum(votes)
cumulative_votes_reversed = np.cumsum(votes[::-1])
x = np.arange(len(states))

plt.figure(figsize=(18, 8))
plt.step(x, cumulative_votes, where='post', color='navy', linewidth=2, marker='o')
x_reversed = np.flip(x)
plt.step(x_reversed, cumulative_votes_reversed, where='post', color='red', linewidth=2, marker='o')

# Threshold for filling
dem_threshold = np.argmax(cumulative_votes >= 227)
plt.fill_between(x[:dem_threshold + 1], cumulative_votes[:dem_threshold + 1], step='post', color='blue', alpha=0.4)

# Adding threshold line and markers
plt.axhline(270, color='purple', linestyle='--', linewidth=2)
plt.text(len(states) / 2, 275, '270 votes to win', horizontalalignment='center', color='purple', fontsize=12)
plt.xlabel('Electoral College by State', fontsize=12)
plt.ylabel('Total Votes', fontsize=12)
plt.title('Cumulative Electoral Votes by State', fontsize=15)
plt.grid(True, linestyle='--', alpha=0.5)
plt.tight_layout()
plt.show()

```

It’s important to note that each pollster employs unique sampling techniques and methodologies that influence their Pollscore and Transparency Score. Pollscore is a measure of a pollster's historical accuracy, reflecting the quality of their sampling methods and data collection. a good score reflecting low propensity to be impacted by errors and biases that can occur in survey sampling. It also accounts for impacts that might bias a survey's score, such as adjustmenting for the differing difficulty of polling certain political races or elements of luck in samples by resampling polls. Transparency Score, on the other hand, measures how openly a pollster shares methodological details like question wording, sampling methods, and weighting procedures. By combining these two metrics into a single star rating, we achieve an excellent overall assessment of a pollster's quality, capturing both their empirical accuracy and their commitment to methodological transparency.
We use this single measurement out of 3 to ensure the integrity of our analysis. Selecting only those polls that originate from reputable pollsters, specifically targeting those with a numeric grade of 2.5 or higher.  This approach is used so that the data reflects a robust and credible representation of voter preferences.

In keeping with this aim, we focused our analysis on polls conducted after September. Early in the election cycle, polls can be significantly influenced by initial campaign events, such as the shocks from the first debates or unexpected changes in the candidate lineup, for example Biden dropping out as a candidate on July 21st. These factors can introduce biases and fluctuations that do not necessarily reflect the enduring preferences of the electorate. By selecting polls from after September, we capture data from a period when voter opinions are more stable and better represent the current state of the race.

```{python}
#| echo: false

# Load the data (replace with your CSV file path)
df = pd.read_csv('../data/02-analysis_data/merged_swing_state_data.csv')

# Group by state and calculate summary statistics
summary_stats = df.groupby('state').agg(
    numeric_grade_mean=('numeric_grade', 'mean'),
    numeric_grade_std=('numeric_grade', 'std'),
    numeric_grade_min=('numeric_grade', 'min'),
    numeric_grade_max=('numeric_grade', 'max'),
    numeric_grade_count=('numeric_grade', 'count'),
    sample_size_mean=('sample_size', 'mean'),
    sample_size_std=('sample_size', 'std'),
    sample_size_min=('sample_size', 'min'),
    sample_size_max=('sample_size', 'max'),
    sample_size_count=('sample_size', 'count')
).reset_index()

# Reshape the summary table to long format
summary_long = pd.melt(
    summary_stats,
    id_vars='state',
    value_vars=[
        'numeric_grade_mean', 'numeric_grade_std', 'numeric_grade_min', 'numeric_grade_max', 'numeric_grade_count',
        'sample_size_mean', 'sample_size_std', 'sample_size_min', 'sample_size_max', 'sample_size_count'
    ],
    var_name='variable_stat', value_name='value'
)

# Split 'variable_stat' into 'variable' and 'stat'
summary_long[['variable', 'stat']] = summary_long['variable_stat'].str.rsplit('_', n=1, expand=True)

# Drop the original column
summary_long.drop('variable_stat', axis=1, inplace=True)

# Pivot the table to get desired output
summary_pivot = summary_long.pivot_table(
    index=['state', 'variable'], columns='stat', values='value', aggfunc='first'
).reset_index()

# Sort for better readability
summary_pivot = summary_pivot.sort_values(by=['state', 'variable'])

summary_pivot[['mean', 'std', 'min', 'max']] = summary_pivot[['mean', 'std', 'min', 'max']].round(1)

# Create a list of rows for tabulate (merging state rows visually)
rows = []
last_state = None
for _, row in summary_pivot.iterrows():
    state = row['state']
    variable = row['variable']
    mean, std, min_, max_, count = row['mean'], row['std'], row['min'], row['max'], row['count']

    # Only show state name on the first row of each group
    if state == last_state:
        state = ''
    else:
        last_state = state

    rows.append([state, variable, mean, std, min_, max_, count])

# Print the table using tabulate
print(tabulate(rows, headers=['State', 'Variable', 'Mean', 'Std', 'Min', 'Max', 'Count'], tablefmt='fancy_grid'))
```

 # 2.2 Variables of Interest
In our logistic regression model predicting the 2024 U.S. Presidential Election outcome, the primary variables of interest are derived from aggregated state-level polling data. The percentage of respondant favouring a candidate become the key variables for prediction. Within the collection of polls in our sample the independent variable `trump_pct` represents the percentage of respondents who indicate support for Donald Trump , and `harris_pct` respectively.
Polls ar taking towards the build up of an election and thus ar taken by repondants at different times. Graph  x shows how the sample of polls are distributed over time in which it is clear that we have a collection of polls from a wide range of periods. Because sentiment changes over time we will adjust the weight that we place on polls that are closer to the election as these will be more indicative of how respondants will be voting on the day.
```{python}
#| echo: false

# Load data (use an accurate path or dataset)
data = pd.read_parquet('../data/02-analysis_data/merged_swing_state_data.parquet')
data['end_date'] = pd.to_datetime(data['end_date'])

plt.figure(figsize=(12, 6))
plt.hist(data['end_date'], bins=20, edgecolor='black', alpha=0.7)
plt.xlabel('Poll End Date')
plt.ylabel('Poll Count')
plt.title('Poll Distribution Over Time')
plt.xticks(rotation=45)
plt.tight_layout()
plt.show()

```

Another aspects of the polls we wish to consider is the population size of each. Larger polls result in a stronger reduction in randomness in the response. The larger the sample size the more the response reflects the true population of the state, this is known as the law of large numbers. In table x we can see the variety within the states polls.
For use in the model the percentages in these polls is weighted for recenecy and size of the population.

@@

 # 4 Results {#sec-results}
 ## Regression results

 ```{Python}
#| echo: false

import pandas as pd
from tabulate import tabulate

# Data setup
data = {
    'State': ['Arizona', 'Georgia', 'Michigan', 'Nevada', 'North Carolina', 'Pennsylvania', 'Wisconsin'],
    'Trump % (Mean)': [49.85, 49.39, 48.57, 47.9, 48.48, 47.68, 47.76],
    'Trump % (SD)': [1.24, 1.78, 2.14, 1.5, 1.41, 1.81, 2.0],
    'Harris % (Mean)': [47.04, 47.66, 47.72, 48.63, 48.58, 48.53, 48.92],
    'Harris % (SD)': [2.17, 2.52, 2.12, 1.19, 1.55, 1.29, 1.72],
    'Percentage Chance of Trump Winning': [86.85, 71.18, 53.09, 21.34, 32.31, 19.67, 15.52]
}

# Create DataFrame
df = pd.DataFrame(data)

# Display as a professional table
table_output = tabulate(
    df, headers='keys', tablefmt='grid', showindex=False,
    floatfmt=(".2f", ".2f", ".2f", ".2f", ".2f", ".2f")
)
print(table_output)

```

in figure xx we cans see th results from the model described.

In Arizona and Georgia, the model predicts the strongest likelihood of a Trump victory, with a mean polling percentage showing Trump over 2 percentage points ahead in Arizona. In Georgia, Trump's mean polling percentage is 49.39% against Harris's 47.66%.The data is thus resulting in an strong 86.85% and and 71.18% likelyhood in Trump winning in Arizona and Georgia respectively.
Trump is also predicted to win in Michigan, alhough by smaller margins as shown in figure xx. As a rssult the model predicts trump winning with only a 53% chance. With Arizona, Georgia and Michigan Trump is predicted a total of 42 of the 93 swing state votes.

```{python}
#| echo: false

# Data setup (can adjust based on model results)
data = {
    'State': ['Arizona', 'Georgia', 'Michigan', 'Nevada', 'North Carolina', 'Pennsylvania', 'Wisconsin'],
    'Trump % (Mean)': [49.85, 49.39, 48.57, 47.9, 48.48, 47.68, 47.76],
    'Trump % (SD)': [1.24, 1.78, 2.14, 1.5, 1.41, 1.81, 2.0],
    'Harris % (Mean)': [47.04, 47.66, 47.72, 48.63, 48.58, 48.53, 48.92],
    'Harris % (SD)': [2.17, 2.52, 2.12, 1.19, 1.55, 1.29, 1.72],
    'Trump Win Probability (%)': [86.85, 71.18, 53.09, 21.34, 32.31, 19.67, 15.52]
}
df = pd.DataFrame(data)
df['Distance from 50%'] = df['Trump Win Probability (%)'] - 50

# Centered bar chart
fig, ax = plt.subplots(figsize=(10, 6))
colors = ['red' if x > 0 else 'blue' for x in df['Distance from 50%']]
ax.barh(df['State'], df['Distance from 50%'], color=colors)
ax.axvline(0, color='black')
ax.set_xlabel('Distance from 50% (Trump Winning Probability)')
ax.set_title('Probability of Trump Win, Centered from 50%')
plt.tight_layout()
plt.show()

```

Kamala is predicted to win he remaining of th states and is givn fairly high chances of locking these up. despite only winning Nevada, North Carolina and Pennsilvania by under a percntag point in the model Kamalas probablity of winning thos eelectoral votes is vry high. This is due to the small standard dviations shown in table x making he prdicted rul of the states leaning democrat being mor likely.
Wisconson is shown to be her srongest and most likley victory of the swing states. 
Overall this givs Harris 51 of the 93 swing state votes his results in a win for the democrats and the first female US president!

## Trump overall probability
-- Taking he probabilities from the model above we can look at the outcomes of the 207 possible combinations of wins and loses for Trump in these swing states adn deduce his overall probability of reaching 270 electoral votes. Basd on the modles probabilities his current expected votes is 41.05, below the threshold of 51 he needs based on our assumptions.

Seperating the outcomes into those that give trump the required electoral college votes, we can obtain  probability that trump ceizes a second term.

Using a probaboiility theory, and assuming independence among the states, in our approach. When accounting for all possible state outcome combinations, the model calculates that Trump has approximately a 22.5% chance of securing at least 51 electoral votes from these key swing states. This probability aggregates the likelihoods across all combinations where the sum of electoral votes meets or exceeds the 51-vote threshold. The analysis reveals that despite high individual probabilities in certain states, the overall chance of Trump amassing sufficient electoral votes from the swing states is less than one in four.

@@

An example of the survey can be found here:
https://docs.google.com/forms/d/e/1FAIpQLSc5LYCAd0OmiIc-LHRFUVgVGuSMKzayVWZB7VOrsrDOe5742w/viewform?vc=0&c=0&w=1&flr=0

# appndix B 
say what kind of ampling. 
reference rsearch on wha the sampling mthod is

## what is included in the poll?

## Methodology and why?

# References
